Thread Management for High Performance Database Systems - Design and Implementation \\{\large Technical Report

### 2018-08-17 Presentations/Paper/paper_thread_pool.tex
More references and text adaption

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{multicol}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{url}

@@

\definecolor{mGreen}{rgb}{0,0.6,0}
\definecolor{mGray}{rgb}{0.5,0.5,0.5}
\definecolor{mPurple}{rgb}{0.58,0,0.82}
\definecolor{backgroundColour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{CStyle}{
	backgroundcolor=\color{backgroundColour},   
	commentstyle=\color{mGreen},
	keywordstyle=\color{blue},
	numberstyle=\tiny\color{mGray},
	stringstyle=\color{mPurple},
	basicstyle=\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	captionpos=b,                    
	keepspaces=true,                 
	numbers=left,                    
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
	language=C,
	morekeywords={*,thread\_pool\_t, thread\_task\_t}
}

@@

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB requires
        spawning many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect we implemented a thread pool to be
        used in GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel,
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration,
        depending on the use case, has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given requirements and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

\begin{IEEEkeywords}
Thread Pool, GeckoDB, BOLSTER
\end{IEEEkeywords}

\section{Introduction}
Since the amount of data that is stored and processed by modern database systems is growing fast, sequential data processing as only possibility is inconceivable. Applications have to process data in parallel to reach sufficient throughput to fulfil appropriate requirements.

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. In this paper, we focus on multi threading by implementing a thread pool for the graph database system GeckoDB\footnote{GeckoDB source code repository: \url{https://github.com/geckodb}}. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. BOLSTER has similarities to Intel TBB~\cite{books/daglib/0018624} but is handwritten from scratch to fit perfectly into GeckoDBs storage engine written in C11~\cite{pinneckegridtables}. In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, a couple of drawbacks arise from this implementation.
First of all, the creation of threads comes along with overhead like stack initialisation and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is 	time-consuming.

To overcome these drawbacks, we integrate an optimised thread pool in BOLSTER. Along with the implementation, we measure the performance of the primitives to determine the thread pool overhead. Additionally, we measure metrics like \emph{waiting and busy time} of threads to evaluate correct thread pool sizes for the considered use cases.

@@

\begin{figure*}[htbp]
	\includegraphics[width=1.0\textwidth]{img/pool_structure.png}
	\caption{Design of the thread pool system}
	\label{fig0}
\end{figure*}

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{img/pool_queue.png}
	\caption{Workflow of task enqueueing and execution in the thread pool}
	\label{fig1}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=0.4\textwidth]{img/waitingconcept.png}
	\caption{Concept of waiting for tasks using a slot-map}
	\label{fig2}
\end{figure}

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle]
	size_t ind = 0;
	// linear search, compact slot usage
	for(; ind < pool->task_state_capacity && pool->task_group_states[ind].task_count; ++ind);
	// distributed over different cache lines
	for(; pool->task_group_states[ind].task_count; ind = (ind + 8) % MAX_NUM_TASKS);
	\end{lstlisting}
	\caption{Different implementations for slot searching}
	\label{fig10}
\end{figure}

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle]
		void cmp_baseline(int numThreads)
		{
			struct timespec begin, end;
			clock_gettime(CLOCK_MONOTONIC, &begin);

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle]
		void cmp_pool(int numThreads)
		{
			thread_pool_t* pool = 
			thread_pool_create(numThreads, 1);

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{img/slot_distr.png}
	\caption{Utilization \% with different slot distribution strategies}
	\label{fig8}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{img/lock_free.png}
	\caption{Comparison of Utilization \% of the lock-free queue versus mutex based heap}
	\label{fig9}
\end{figure}

@@

\section{Acknowledgements}
This work was partially funded by the DFG (grant no.: SA 465/50-1).

\bibliographystyle{splncs}
\bibliography{paper_thread_pool}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{multicol}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{balance}
\usepackage{listings}
\usepackage{url}

@@

\definecolor{mGreen}{rgb}{0,0.6,0}
\definecolor{mGray}{rgb}{0.5,0.5,0.5}
\definecolor{mPurple}{rgb}{0.58,0,0.82}
\definecolor{backgroundColour}{rgb}{1,1,1}

\lstdefinestyle{CStyle}{
	backgroundcolor=\color{backgroundColour},   
	commentstyle=\color{mGreen},
	keywordstyle=\color{blue},
	numberstyle=\tiny\color{mGray}\fontsize{4}{4}\selectfont\ttfamily,
	stringstyle=\color{mPurple},
	basicstyle=\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	captionpos=b,                    
	keepspaces=true,                 
	numbers=left,                    
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
	language=C,
	basicstyle=\fontsize{7}{7}\selectfont\ttfamily,
	morekeywords={*,thread\_pool\_t, thread\_task\_t}
}

@@

\begin{abstract}
        Processing tasks in parallel is used in nearly all applications to keep
        up with the requirements of modern software systems. However, the
        current implementation of parallel processing in GeckoDB, a graph database system developed in our group, requires
        to spawn many short-lived threads that execute a single task and then
        terminate. This creates a lot of overhead since the threads are not
        being reused. To counter this effect we implemented a thread pool to be
        used in GeckoDB to reduce the additional time required
        to create and close thousands of threads. In this paper, we show our
        implementation of a thread pool to process independent tasks in parallel,
        including the waiting for a set of processed tasks. We also compare the
        thread pool implementation against the current one. Additionally, we show that the task and thread pool configuration,
        depending on the use case, has a high impact on the thread
        pool performance. At the end of our evaluation, we show that the implementation 
        fulfils all given requirements and generally reduce
        overhead compared to creating and spawning threads individually.
\end{abstract}

\begin{IEEEkeywords}
Database Management System, Architecture of Parallel Processing Systems, Thread and Task Management
\end{IEEEkeywords}

\section{Introduction}
Since the amount of data that is stored and processed by modern database systems is growing fast, sequential data processing as only possibility is inconceivable~\cite{Alonso:2013:HKS:2510649.2511162}. Hence, applications and systems have to process data in parallel to reach sufficient throughput to fulfil appropriate requirements~\cite{DBLP:series/synthesis/2017Ailamaki, 10.1007/978-3-319-72413-3_8, 10.5120/ijca2016912260} which in turn affects the architectural design and query processing of modern database systems~\cite{agrawal2017many, Makreshanski2017, schuh2016experimental, psaroudakis2016adaptive, psaroudakis2013task}.

The growing number of cores for both traditional CPUs as well as trending co-processors like GPUs require to re-evaluate core design descicion in data-intense systems. For instance, Wu et al. examined the scalability of multi-version concurrency control (MVCC) for a growning number of threads in  2017\cite{wu2017empirical}. They concluded that multi-threading in particular and modern hardware in general both promise notable performance gains for high-performance database systems using MVCC. However, they also observed that there is no clear winner combination for several concurrency control protocol, version storage, and garbage collection as well as the index management strategies in MVCC.

For this paper, we focus on managing a far more lower level: general parallel data processing. In particular, with this paper we provide insights into our thread managing strategy in GeckoDB/BOOLSTER, its design and its implementation.

Parallel data processing can be achieved by different approaches, like instruction and data parallelism or multi threading. We focus on multi threading by implementing a thread pool for the graph database system GeckoDB\footnote{GeckoDB source code repository: \url{https://github.com/geckodb}}. The thread pool will be integrated into BOLSTER, a high performance library for parallel execution of primitives like for or filter on large data sets. BOLSTER has similarities to Intel TBB~\cite{books/daglib/0018624}, a C++ template library for parallel programming, but BOLSTER is written from scratch to perfectly fit into GeckoDBs specialized storage engine and vectorized query engine, both written in C11~\cite{pinneckegridtables}.

In the current implementation, BOLSTER creates a fix number of threads for each call of a primitive. This approach is called \emph{thread-per-request}. Since many primitives are executed at the same time, a couple of drawbacks arise from this implementation.
First of all, the creation of threads comes along with overhead like stack initialization and memory allocation. Secondly, creating a huge number of threads simultaneously may lead to large context switch overhead of the scheduler. Additionally, debugging and profiling applications that create many threads during runtime is time consuming.

To overcome these drawbacks, we integrate an optimized thread pool in BOLSTER. Along with the implementation, we measure the performance of the primitives to determine the thread pool overhead. Additionally, we measure metrics like \emph{waiting and busy time} of threads to evaluate correct thread pool sizes for the considered use cases.

@@

\begin{figure*}[htbp]
	\centering \includegraphics[width=0.7\textwidth]{img/pool_structure.png}
	\caption{Design of the thread pool system}
	\label{fig0}
\end{figure*}

@@

\begin{figure}
	\centering\includegraphics[width=0.4\textwidth]{img/pool_queue.png}
	\caption{Workflow of task enqueueing and execution in the thread pool}
	\label{fig1}
\end{figure}

@@

\begin{figure}
\centering	\includegraphics[width=0.25\textwidth]{img/waitingconcept.png}
	\caption{Concept of waiting for tasks using a slot-map}
	\label{fig2}
\end{figure}

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle,numbers=none]
	size_t ind = 0;
	// linear search, compact slot usage
	for(; ind < pool->task_state_capacity && pool->task_group_states[ind].task_count; ++ind);
	// distributed over different cache lines
	for(; pool->task_group_states[ind].task_count; ind = (ind + 8) % MAX_NUM_TASKS);
	\end{lstlisting}
	\caption{Different implementations for slot searching}
	\label{fig10}
\end{figure}

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle,numbers=none]
		void cmp_baseline(int numThreads)
		{
			struct timespec begin, end;
			clock_gettime(CLOCK_MONOTONIC, &begin);

@@

\begin{figure}
	\begin{lstlisting}[style=CStyle,numbers=none]
		void cmp_pool(int numThreads)
		{
			thread_pool_t* pool = 
			thread_pool_create(numThreads, 1);

@@

\begin{figure}
	\centering\includegraphics[width=0.4\textwidth]{img/slot_distr.png}
	\caption{Utilization \% with different slot distribution strategies}
	\label{fig8}
\end{figure}

\begin{figure}
	\centering\includegraphics[width=0.4\textwidth]{img/lock_free.png}
	\caption{Comparison of Utilization \% of the lock-free queue versus mutex based heap}
	\label{fig9}
\end{figure}

@@

\section{Acknowledgements}
This work was partially funded by the DFG (grant no.: SA 465/50-1). Also we would like to thank Sebastian Bre\ss, Thomas Leich, David Broneske, and Gabriel Campero Durand.

\bibliographystyle{splncs}
\balance
\bibliography{paper_thread_pool}
